### 2010-04-01 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: WeightWatch section 3, A universe of inductive types

In this section, we describe an implementation of inductive types, as
we know them in ML-like languages. This choice is motivated by
pedagogical considerations: by describing the layman datatypes in our
setting, we build on our intuition of datatypes we are familiar
with. In particular, we do not aim at describing a model for generics
in simply-typed functional language. Our goal is and remains to
discuss datatypes as we find them in dependently-typed languages.

Our proposal is inspired by, but less involved than, Dybjer and
Setzer's closed formulation of
induction-recursion~\cite{dybjer:general-ir, dybjer:axiom-ir,
  dybjer:ir-initial-algebra, dybjer:iir}. For inductive datatypes, the
full power of induction-recursion is not necessary. Therefore, we
adopt a stripped-down version of it. As we broaden the scope of our
universe of datatypes, our presentation will diverge from
induction-recursion. We shall compare both approaches in
Section~\ref{sec:discussion}.

@@

In dependently-typed languages, $\Sigma$-types can be interpreted in
two different ways. This duality is actually reflected in the notation
we can find in the literature, depending on the sensibility of the
author.  The notation $\blue{\Sigma}_{\V{x} : \V{A}} (\V{B}\: \V{x})$
stresses the interpretation that $\Sigma$-types are a generalisation
of sums over arbitrary arities.  Where simply-typed languages have
finite sums, dependently-typed languages have sums indexed by any set.

On the other hand, our choice of notation $\SIGMA{\V{x}}{\V{A}} (\V{B}\:\V{x})$
emphasises that $\Sigma$-types can be read as a generalisation of products,
where the second component can depend on the first one. When simply-typed
languages pack data into non-dependent tuples, dependently-typed languages
tolerate this non-dependent usage. However, they also give the rather novel
ability for data to influence types of further data.

@@

In the simply-typed world, datatype definitions are often referred to
as \emph{sum-of-product}. A datatype is defined by a finite sum -- a
choice -- of constructors, each of them is composed of a product -- a
tuple -- of arguments. This form governs the grammar of datatype
definition in languages such as Haskell or ML.

To model these datatypes, we simply need to capture this grammar in
our dependently-typed setting. Here, the notion of sum-of-product
naturally translates into \emph{sigmas-of-sigmas}. Beyond this
intuition, it should be clear that a direct encoding of datatypes
through raw $\Sigma$-types is not a viable option. Indeed, just as for
finite sets, encoding throws away information when we crucially need
it. Moreover, to program with these datatypes, one would need an
induction principle. An encoding will be unable to deliver such
operation~\cite{geuvers:induction-not-derivable}.

@@

While sigmas-of-sigmas form a convenient \emph{semantics} for
datatypes, we have seen that it makes a poor \emph{syntax}. To
decouple these two notions, we use a standard technique in
dependently-typed programming: we construct an ad-hoc universe. This
technique dates back to Martin-L\"of definition of type
theory~\cite{martin-lof:itt}. Since then, it has been fruitful as a
programming technique~\cite{benke:universe-generic-prog,
  oury:power-of-pi} \note{Any idea of other papers?}.  We refer the
reader to the Agda tutorial~\cite{norell:agda-tutorial} for a
pedagogical presentation of universe construction.  \note{Is that
  really a motivating motivation for the usage of universe
  construction?}  \note{pwm: Surely ($\EnumU$, $\EnumT$) is a useful
  example of universes too.  Even if it doesn't show off their full
  power?}

The key idea behind universe construction is our ability to make names
by defining new types. These names are called \emph{codes}. By
defining a set of codes, we define the syntax of a language. However,
as such, a system of codes is useless as it lacks a semantics. Instead
of equipping the universe of codes with some computational behaviour,
we pragmatically chose to \emph{interpret} these codes back into the
standard type theory. Hence, codes act only as labels, while the type
theory provides the computational machinery. \note{Is that a good
  description?}

Codes being simple labels, we have the ability to inspect them, hence
regaining structural information. Being able to inspect them, we are
therefore free to compute over them: deriving new codes from previous
ones, or even new functions on them. In several occasions, we will
have the opportunity to witness the power of universes.

@@

We propose to embed inductive types as a universe in our dependent
type theory, the universe of \emph{descriptions}
(Figure~\ref{fig:desc_universe}). As expected, the code of this
universe mimics the standard grammar of datatypes definitions in
simply-typed languages. Hence, we have a $\DSigma{\!}{\!}$ code,
interpreted as a $\Sigma$-type, to build the
sigmas-of-sigmas. Descriptions are terminated by $\DUnit$, which
contains no useful payload. The functoriality of the datatypes is
introduced by $\DIndx{\!}$. When we tie the knot with a fix-point, the
hole left open by $\DIndx{\!}$ will be turned into a standard
recursive argument. We notice that this functoriality appears in the
type of $\descop{\V{D}}{\!}$ itself, for a given datatype definition
$\V{D}$. This corresponds to the object part of the functor described
by $\V{D}$.

@@

To give some intuition on this universe of descriptions, we now turn
to some examples. For obvious pedagogical reasons, we will manually
build these descriptions. However, it should be clear that, in
practice, these definitions can be automatically constructed from an
ML-like $\data$ definition.

@@

Our first example is the natural numbers, or rather its pattern
functor. Our code is presented in the high-level expression language
of Section~\ref{sec:type-propagation}. The translation back to the raw
terms is laborious but should not pose any difficulty. The code is the
following:

@@

Let us explain its construction. First, we use $\DSigma{\!}{\!}$ to
give a choice between the $\NatZero$ and $\NatSuc{\!}$
constructors. In the $\NatZero$ case, we reach the end of the
description: there is no useful payload in that case. In the
$\NatSuc{\!}$ case, we leave one hole open for the recursive argument,
and we close the description.

In a more synthetic notation, we have described the following functor:

@@

With a small change to the definition of $\NatD$, we obtain the
pattern functor for lists:

@@

The modification consists in turning the $\NatSuc{\!}$ constructor
into a proper $\ListCons{\!}{\!}$ taking an argument in $\V{X}$
followed by a recursive argument. In this case, we use
$\DSigma{\!}{\!}$ in its product interpretation: we pack an element of
$\V{X}$ together with the recursive argument. Easily, one sees that
this code actually describes the following functor:

@@

Again, this is a clear reminiscence of the sum-of-product style:
we have a finite sum of constructors. Of course, every description can
be expressed in this style, using a singleton as single constructor if
necessary. A tagged description can trivially be turned into a
full-blown description. We write $\toDesc{\V{TD}}$ for the description computed
from the tagged description $\V{TD}$.

@@

For convenience, we have taken for granted the existence of $\Desc$,
as presented in Figure~\ref{fig:desc_universe}. In particular, we have
considered its codes, $\DSigma{\!}{\!}$, $\DIndx{\!}$, and $\DUnit$ as
type formers, extending the basic type theory. Hence, in an
implementation of this type theory, we need to extended it with these
constructors and their typing rules. Although it makes no conceptual
difference, the code of description should rather be read as a
specification. We promise the existence of such objects, satisfying
the typing rules. It will be the subject of
Section~\ref{sec:desc-levitate} to fulfil this promise, by actually
implementing the specification.

@@

So far, we have used our universe of descriptions to build
functors. \note{ Strictly positive types anyone? } To illustrate its
usage, we have implemented the pattern functors of natural numbers,
lists, and binary trees. However, the class of functors expressible
with descriptions enjoys another property: they all admit a
fix-point. Indeed, descriptions are interpreted into \emph{polynomial
  functors}~\cite{who?}\note{who is the relevant source for polynomial
  functors?}, as they are only composed of $\Sigma$-types, the unit
type, and variables.

We construct the fix-point by \emph{tying the knot} of the
interpretation: the holes left open by the functor are filled by its
own recursive definition:

@@

Whereas this definition is amply sufficient in the simply-typed world,
it comes short of its promises in the dependently-typed realm. Indeed,
this type is not dependent, being restricted to a notion of
\emph{iteration}. As this operator will be the corner stone of any
inductive definition, we are better off sharpening it before use.
Following \citet{benke:universe-generic-prog}, we adopt the following
definition:

\[
\begin{array}{lcll}
\F{induction} & : & \multicolumn{2}{l}{\PITEL{\V{D}}{\Desc}}                   \\
              &   & \multicolumn{2}{l}{\PITEL{\V{x}}{\Mu{\V{D}}}}                   \\
              &   & \multicolumn{2}{l}{\PITEL{\V{P}}{\Mu{\V{D}} \To \Set}}         \\
              &   & \PI{\V{m}}{& \PI{\V{xs}}{\descop{\V{D}}{(\Mu{\V{D}})}}              \\
              &   &           & \All{\V{D}}{(\Mu{\V{D}})}{\V{P}}{\V{xs}} \To \V{P} (\Con{\V{xs}})} \\
              &   & \V{P} \V{x}
\end{array}
\]

Where, intuitively, $\All{\V{D}}{\V{X}}{\V{P}}{\V{x}}$ states that
$\Bhab{\V{P}}{\V{X} \To \Set}$ holds everywhere in $\V{x}$. Its formal
definition is presented in Figure~\ref{fig:all-predicates}.  We refer
the reader to our technical report for the implementation of
$\F{induction}$. Having this dependent catamorphism, we are empowered
with the ability to $\emph{compute}$ over descriptions. Note that this
induction principle is the first manifestation of a generic operation
over descriptions. Thanks to $\F{induction}$, any datatype we define
automatically comes equipped with an induction principle.

@@

Whereas simply-typed languages feature a convenient grammar for
defining datatypes and manipulating them, our proposal could be seen
as a step backward. In our setting, the developer would have to write
a description instead of a sugared datatype definition. Then, to
recognise an element of $\Nat$ for instance, she would have to
identify $\Mu{}$ of some coded description as indeed the fix-point of
the pattern functor of natural numbers. Finally, she would have to
write codes instead of datatype constructors.

As for finite sets previously, these symptoms clearly indicate that we
suffer from too much type information. Just as for finite sets, the
cure lies in rationalising this information. Let us first tackle the
issue of datatype declaration. The grammar of datatypes we are used
to can be straightforwardly translated to the sigmas-of-sigmas
paradigm. Hence, a datatype definition can be turned into the
corresponding description. Then, at the level of definitions, we
associate the name of the datatype to the fix-point of the
description. The induction principle as well as other generic
operations follow for free.

Doing so, we seal datatype definition behind a convenient
abstraction. We take advantage of this abstraction barrier to address
the second issue. Namely, the developer would rather be presented a
name than a fix-point of a pattern functor. Surprisingly, it is
sufficient to \emph{label} the $\Mu{}$ constructor with the
user-defined name. When pretty-printing this object, we just expose
the label, instead of the definition in full.

Regaining the ability to write datatype constructors is simply a
matter of type propagation. We therefore extend the type-checking
framework of Figure~\ref{fig:type-checking} with the following
inference rule:

@@

Where $\vec{\V{e}}$ denotes a telescope of arguments applied to the
constructor $\V{C}$. For the sake of readability, we write function
application in the small function space as $\V{f}\: \V{C}$. This
corresponds to a call to $\F{switch}$. Also, we note
$\widetilde{\V{e}}$ the right-nested tuple terminated by $\Void$
corresponding to the curried telescope $\vec{\V{e}}$.

@@

This concludes our presentation of the universe of descriptions. In
this section, we have seen how datatypes from the simply-typed world
can be expressed in our dependently-typed setting. By relying on a
universe construction technique, we benefit from the ability to
inspect and manipulate these codes. This opens some appealing
opportunities for generic programming. Finally, we have demonstrated
that programming with codes is practical. To this end, we have
extended the type propagation machinery. In this setting, we can
define, interact, and build datatypes as transparently as in the
simply-typed world.

>>>>

In this section, we describe an implementation of inductive types, as
we know them in ML-like languages. By working with datatypes we are
familiar with, we rely on our intuition on these objects. However, our
goal remains to discuss datatypes as we find them in dependently-typed
languages.

Our proposal is inspired by Dybjer and Setzer's closed formulation of
induction-recursion~\cite{dybjer:axiom-ir}. For inductive datatypes, the
full power of induction-recursion is not necessary. Therefore, we
adopt a stripped-down version of it.

@@

In dependently-typed languages, $\Sigma$-types can be interpreted in
two different ways. This duality is actually reflected in the notation
we can find in the literature. The notation $\blue{\Sigma}_{\V{x} :
  \V{A}} (\V{B}\: \V{x})$ stresses that $\Sigma$-types are a
generalisation of sums over arbitrary arities.  Where simply-typed
languages have finite sums, dependently-typed languages have sums
indexed by any set.

On the other hand, our choice of notation $\SIGMA{\V{x}}{\V{A}}
(\V{B}\:\V{x})$ emphasises that $\Sigma$-types are a generalisation of
products, where the second component can depend on the first
one. Where simply-typed languages pack data into non-dependent tuples,
we can now build tuples where data influence further types.

@@

In ML-like languages, datatype definitions are called
\emph{sum-of-product}. A datatype is defined by a finite sum of
constructors, each composed by a product of arguments. To embrace
these datatypes, we need to capture this grammar.

With dependent types, the notion of sum-of-product translates into
\emph{sigmas-of-sigmas}. However, a direct encoding of datatypes
through raw $\Sigma$-types is not a viable option. Indeed, to program
with these datatypes, we need an induction principle. An encoding is
unable to deliver such
operation~\cite{geuvers:induction-not-derivable}.

@@

While sigmas-of-sigmas form a convenient \emph{semantics} for
datatypes, we have seen that it makes a poor \emph{syntax}. To
decouple these two notions, we construct an ad-hoc
\emph{universe}. This technique originates in Martin-L\"of definition
of type theory~\cite{martin-lof:itt}. It is a fruitful programming
technique in dependently-typed
languages~\cite{benke:universe-generic-prog, oury:power-of-pi}
\note{Any idea of other papers?}.  \note{Is that really a motivating
  motivation for the usage of universe construction?}  \note{pwm:
  Surely ($\EnumU$, $\EnumT$) is a useful example of universes too.
  Even if it doesn't show off their full power?}

The key idea behind universe construction is our ability to make names
by defining new types. These names are called \emph{codes}. By
defining a set of codes, we define the syntax of a language. To give
it a semantics, we \emph{interpret} these codes back into the type
theory. Hence, codes act as labels, while the type theory provides the
computational machinery. \note{Is that a good description?}

Codes being mere labels, we can inspect them, hence regaining
structural information. We can also compute over them: deriving new
codes from others, or even functions on them.

@@

To embed inductive types in our type theory, we define a universe of
\emph{descriptions} (Figure~\ref{fig:desc_universe}). As expected, the
code of this universe mimics the grammar of inductive types. To build
sigmas-of-sigmas, we define a $\DSigma{\!}{}$ code, interpreted as a
$\Sigma$-type. Descriptions are terminated by $\DUnit$, containing no
payload. To make recursive arguments, we introduce the $\DIndx{\!}$
code. It interprets to a hole that will be closed by the fix-point. We
can also read $\V{D}$ as describing an endofunctor on $\Set$, with
$\descop{\V{D}}{\!}$ being the object part.

@@

To give some intuition on this universe, we now turn to some
examples. For obvious pedagogical reasons, we will manually build
these descriptions. However, it should be clear that, in practice,
these definitions are automatically computed from an ML-like $\data$
definition.

@@

Our first example is the natural numbers, or rather its pattern
functor. Our code is presented in the high-level expression language
of Section~\ref{sec:type-propagation}. The translation back to the raw
terms should not pose any difficulty. The code is the following:

@@

Let us explain its construction. First, we use $\DSigma{\!}{\!}$ to
give a choice between the $\NatZero$ and $\NatSuc{}$ constructors. In
the $\NatZero$ case, we reach the end of the description. In the
$\NatSuc{\!}$ case, we attach a recursive argument and close the
description. In a more synthetic notation, we have described the
following functor:

@@

With a small change to this definition, we obtain the pattern functor
for lists:

@@

The $\NatSuc{\!}$ constructor is turned into a proper
$\ListCons{\!}{\!}$, taking an argument in $\V{X}$ followed by a
recursive argument. Easily, one sees that this code describes the
following functor:

@@

Again, this is a clear reminiscence of the sum-of-product style: we
have a finite sum of constructors. Every description can be expressed
in this style, using a singleton as a constructor if
necessary. Converselly, a tagged description can be turned into a
full-blown description. We denote $\toDesc{\V{TD}}$ the description
computed from the tagged description $\V{TD}$.

@@

For convenience, we have taken for granted the existence of $\Desc$,
as presented in Figure~\ref{fig:desc_universe}. Although it makes no
conceptual difference, the code of description should rather be read
as a specification. We promise the existence of such objects,
satisfying the typing rules. It will be the subject of
Section~\ref{sec:desc-levitate} to actually implement the
specification.

@@

So far, we have used our universe of descriptions to build pattern
functors. Being polynomial functors, they all admit a fix-point. We
construct the fix-point by \emph{tying the knot} of the
interpretation: the holes left open by the functor are filled by its
own recursive definition:

@@

Whereas this definition is amply sufficient in the simply-typed world,
it comes short of its promises in the dependently-typed realm. Indeed,
this type is not dependent, being restricted to a notion of
\emph{iteration}. As this operator will be the corner stone of
inductive definitions, we are better off sharpening it before use.
Following \citet{benke:universe-generic-prog}, we adopt the following
definition:

\[
\begin{array}{lcl}
\F{induction} & : & \PITEL{\V{D}}{\Desc}
                    \PI{\V{P}}{\Mu{\V{D}} \To \Set}         \\
              &   & (\PI{\V{xs}}{\descop{\V{D}}{(\Mu{\V{D}})}}              \\
              &   & \ \All{\V{D}}{(\Mu{\V{D}})}{\V{P}}{\V{xs}} \To \V{P} (\Con{\V{xs}})) \To \\
              &   & \PI{\V{x}}{\Mu{\V{D}}} \V{P} \V{x}
\end{array}
\]

Where, intuitively, $\All{\V{D}}{\V{X}}{\V{P}}{\V{x}}$ states that
$\Bhab{\V{P}}{\V{X} \To \Set}$ holds everywhere in $\V{x}$. Its
definition is presented in Figure~\ref{fig:all-predicates}.  We refer
the reader to our technical report for the implementation of
$\F{induction}$. Note that this induction principle is the first
manifestation of a generic operation over descriptions. Any datatype
we define automatically comes with an induction principle.

@@

Writing codes instead of datatype constructor is indeed unpractical.
However, regaining this ability is simply a matter of type
propagation. We extend the type-checking framework of
Figure~\ref{fig:type-checking} with the following inference rule:

@@

Where $\vec{\V{e}}$ denotes a telescope of arguments applied to the
constructor $\V{C}$. For readability, we write $\V{f}\: \V{C}$ instead
of a call to $\F{switch}$. Also, we note $\widetilde{\V{e}}$ the
right-nested tuple terminated by $\Void$ corresponding to the curried
telescope $\vec{\V{e}}$.

@@

This concludes our presentation of the universe of descriptions. In
this section, we have seen how datatypes from the simply-typed world
can be expressed in a dependently-typed setting. By relying on a
universe, we benefit from the ability to inspect and manipulate
codes. We have shown that programming with codes is practical.
